---
title: "MRM_prep"
author: "Lili Benitez"
date: "2024-09-17"
output: html_document
---
### load libraries 
```{r, warning=FALSE, include= FALSE}
library(tidyverse)
library(ggplot2)
library(CommEcol)
library(ecodist)
```

### Read in specimen and veg data
```{r}
load("../../Dropbox/skyIslands/data/spec_net.Rdata")

net_specimens<-spec.net%>%
  filter(Method=="Net")#filter by netted insects, actually all are netted I guess

veg<-read_csv("../../Dropbox/skyIslands_saved/data/relational/relational/traditional/veg-complete.csv")

sitestats<-read_csv("../../Dropbox/skyIslands/data/sitestats.csv") #NO LONGITUDE???

```
### Now I want to calculate chao dissimliarity for netted bee communities.
```{r, bee chao dissimilarity}
####bee dissimilarity####
bee<-net_specimens%>%
  filter(Order=="Hymenoptera")%>% #filter all hymenoptera, goes down to 10,783
  filter(!Family=="Sphecidae",!Family=="Braconidae",!Family=="Vespidae") #filter out wasps, goes down to 9,958 observations

bee<-bee%>%
  select(Site, GenusSpecies, Year, SampleRound, State, Meadow) #don't need all the other info
bee_n<-bee%>%
  group_by(Site, Year,SampleRound)%>%
  count(GenusSpecies)%>%
filter(!(Site=="CC"))%>%# CC is just a non-site site
  mutate(SiteYear=paste0(Site, Year))
   #filter(!(SiteYear=="PL2017")) #veg data was messed up here, plus there was only one sample round before fire
bee_all<-data.frame( plots = bee_n$SiteYear, species =bee_n$GenusSpecies,
                      freq = bee_n$n, stringsAsFactors=FALSE)
mean_bee<-crosstab(plots, species, freq, data=bee_all, type="mean")#average species frequencies
bee.dis<-dis.chao(mean_bee, index="jaccard", version='prob') #I think the probability version was better than the rare one for some reason? 
bee.dis<-as.matrix(bee.dis) #at the moment, all site/year combos, diagonal is all zeros which is good

#write.csv(net.bee.chao, "net.bee.prob.csv", row.names=FALSE)
```

### Now we are doing the same thing for the syrphids
```{r, syrph chao dissimilarity}
syrph<-net_specimens%>%
  filter(Family=="Syrphidae") #filter all syrphids, 2,565 observations
syrph<-syrph%>%
  select(Site, GenusSpecies, Year, SampleRound, State, Meadow) #don't need all the other info
syrph_n<-syrph%>%
  group_by(Site, Year,SampleRound)%>%
  count(GenusSpecies)%>%
filter(!(Site=="CC"))%>%# CC is just a non-site site
  mutate(SiteYear=paste0(Site, Year))
  # filter(!(SiteYear=="PL2017")) #veg data was messed up here, plus there was only one sample round before fire
syrph_all<-data.frame( plots = syrph_n$SiteYear, species =syrph_n$GenusSpecies,
                      freq = syrph_n$n, stringsAsFactors=FALSE)
mean_syrph<-crosstab(plots, species, freq, data=syrph_all, type="mean")#average species frequencies
syrph.dis<-dis.chao(mean_syrph, index="jaccard", version='prob') #I think the probability version was better than the rare one for some reason? 
syrph.dis<-as.matrix(syrph.dis) #there were no syrphs at UK, diagonal is all zeros which is good

```

### Now we want to calculate the community dissimilarity for the flowering communities. I originally did this with the bloom dataset, but now I am going to use the veg quadrat data instead, since it includes 2012. Going to be pulling the veg quadrat summary dataset straight from 4_specimenprep so I don't re-do work lol. 

```{r}
veg_quad<-read_csv("veg.bloom.quad.sp.csv")
sum_quad<-veg_quad%>%
  group_by(Site, Year, SampleRound, PlantGenusSpecies)%>%
  mutate(sum_abundance_site_SR=sum(FloweringPlantAbundance), sum_blooms_site_SR=sum(FloralAbundance)) #there are some issues with 2012- abundance is likely actually bloom counts

#meadow.size<-spec.net%>%
#select(Site, Area)%>%
 # unique()
#mean_quad<-left_join(mean_quad, meadow.size, by="Site")
#mean_quad<-mean_quad%>%
 # group_by(Site, Year, PlantGenusSpecies)%>%
 # mutate(mean_abundance_per_year=mean(sum_abundance_site_SR))

#scaled_abundance<-mean_quad%>%
 # mutate(scaled_abundance=mean_abundance_per_year*Area) #if we need to scale up to the meadow size (actually need to fix this calculation since I summed the quadrats instead of averaging)

sum_quad<-sum_quad%>%
   mutate(SiteYear=paste0(Site, Year))%>%
  filter(!(SiteYear=="RP2018"))%>% #no specimen data for RP in 2018???
 # filter(!(SiteYear=="PL2017"))%>% #veg data was messed up here, plus there was only one sample round before fire
  select(SiteYear, PlantGenusSpecies, Year,sum_abundance_site_SR, sum_blooms_site_SR, SampleRound)

#veg_all<-data.frame( plots = sum_quad$SiteYear, species =sum_quad$PlantGenusSpecies,
#                      freq = sum_quad$sum_abundance_site_SR, sampleR=sum_quad$SampleRound, stringsAsFactors=FALSE)%>% #this takes the frequencies of each species at the quadrat level and makes it into a dataframe for 'crosstab'
 # unique()
#mean_veg<-crosstab(plots, species, freq, data=veg_all, type="mean")#average species frequencies across sample round, transpose so species are the columns and siteyears are the rows

bloom_all<-data.frame( plots = sum_quad$SiteYear, species =sum_quad$PlantGenusSpecies,
                      freq = sum_quad$sum_blooms_site_SR, sampleR=sum_quad$SampleRound, stringsAsFactors=FALSE)%>% #this takes the frequencies of each species at the quadrat level and makes it into a dataframe for 'crosstab'
  unique()
mean_bloom<-crosstab(plots, species, freq, data=bloom_all, type="mean")#average species frequencies across sample round, transpose so species are the columns and siteyears are the rows

#veg.dis<-dis.chao(mean_veg, index="jaccard", version='prob') #calculates chao dissimilarity across the siteyears
#veg.dis<-as.matrix(veg.dis)# looks good

bloom.dis<-dis.chao(mean_bloom, index="jaccard", version='prob') #calculates chao dissimilarity across the siteyears
bloom.dis<-as.matrix(bloom.dis)# looks good
##### need to make a separate matrix for syrphs since there were none for UK
sum_quad_syrph<-sum_quad%>%
   mutate(SiteYear=paste0(Site, Year))%>%
  filter(!(SiteYear=="RP2018"))%>% #no specimen data for RP in 2018???
#  filter(!(SiteYear=="PL2017"))%>%#veg data was messed up here, plus there was only one sample round before fire
  filter(!(SiteYear=="UK2017"))%>% 
  select(SiteYear, PlantGenusSpecies, Year, sum_abundance_site_SR,sum_blooms_site_SR, SampleRound)

#veg_syrph<-data.frame( plots = sum_quad_syrph$SiteYear, species = sum_quad_syrph$PlantGenusSpecies, freq = sum_quad_syrph$sum_abundance_site_SR, sampleR=sum_quad_syrph$SampleRound, stringsAsFactors=FALSE)%>% #this takes the frequencies of each species at the quadrat level and makes it into a dataframe for 'crosstab'
 # unique()
#mean_veg_syrph<-crosstab(plots, species, freq, data=veg_syrph, type="mean")#average species frequencies across sample round, transpose so species are the columns and siteyears are the rows

#veg.dis.syrph<-dis.chao(mean_veg_syrph, index="jaccard", version='prob') #calculates chao dissimilarity across the siteyears

#veg.dis.syrph<-as.matrix(veg.dis.syrph)#looks good

bloom_syrph<-data.frame( plots = sum_quad_syrph$SiteYear, species = sum_quad_syrph$PlantGenusSpecies, freq = sum_quad_syrph$sum_blooms_site_SR, sampleR=sum_quad_syrph$SampleRound, stringsAsFactors=FALSE)%>% #this takes the frequencies of each species at the quadrat level and makes it into a dataframe for 'crosstab'
  unique()
mean_bloom_syrph<-crosstab(plots, species, freq, data=bloom_syrph, type="mean")#average species frequencies across sample round, transpose so species are the columns and siteyears are the rows

bloom.dis.syrph<-dis.chao(mean_bloom_syrph, index="jaccard", version='prob') #calculates chao dissimilarity across the siteyears

bloom.dis.syrph<-as.matrix(bloom.dis.syrph)#looks good

```

### we now want to calculate the geographic distances between sites
```{r}
#need some dataframes of the site/years that are for each matrix
bee.sites<-data.frame(Site=sum_quad$Site, SiteYear=sum_quad$SiteYear)%>%
  unique()
syr.sites<-data.frame(Site=sum_quad$Site, SiteYear=sum_quad$SiteYear)%>%
  unique()
syr.sites<-syr.sites%>%filter(!(SiteYear=="UK2017"))
 
library("geosphere")
lat_long<-spec.net%>%
  select(Long, Lat, Site)%>%
  unique()
####make geographic distance matrix for bees
lat_long_bee<-left_join(bee.sites, lat_long, by="Site")%>%arrange(Site)

geo_bee<-lat_long_bee%>%
  select(Long, Lat)
geo.dis_bee <- distm(geo_bee, fun = distGeo)
rownames(geo.dis_bee)<-lat_long_bee$SiteYear
colnames(geo.dis_bee)<-lat_long$SiteYear
geo.bee.km<-geo.dis_bee/1000# Apply distm function
colnames(geo.bee.km)<-rownames(geo.dis_bee)

#write.csv(geo.km, "geo.dist.km.csv", row.names = FALSE)

####make geographic distance matrix for syrphids

lat_long_syrph<-left_join(syr.sites, lat_long, by="Site")%>%arrange(Site)

geo_syrph<-lat_long_syrph%>%
  select(Long, Lat)
geo.dis_syrph <- distm(geo_syrph, fun = distGeo)
rownames(geo.dis_syrph)<-lat_long_syrph$SiteYear
colnames(geo.dis_syrph)<-rownames(geo.dis_syrph)
geo.syrph.km<-geo.dis_syrph/1000#
```

###precipitation matrices: using just annual rainfall, although this could change later on
```{r}
precip<-read_csv("EnvironmentalData/precipitation_all.csv")%>%
  unique()%>%
  mutate(SiteYear=paste0(Site, Year))%>%
  select(-Site, -Year, -Lat, -Long)
precip_bee<-left_join(bee.sites, precip, by="SiteYear")%>%
  unique()%>%
  arrange(Site) #make sure it is in alphabetical order
precip_syrph<-left_join(syr.sites, precip, by="SiteYear")%>%
  unique()%>%
  arrange(Site)

precip.bee.dis<-dist(precip_bee,method = "maximum",diag=FALSE)
precip.bee.dis<-as.matrix(precip.bee.dis)#check to make sure it matches the other matrices
rownames(precip.bee.dis)<-precip_bee$SiteYear
colnames(precip.bee.dis)<-precip_bee$SiteYear

#do the same for syrphids
precip.syrph.dis<-dist(precip_syrph,method = "maximum",diag=FALSE)
precip.syrph.dis<-as.matrix(precip.syrph.dis)#check to make sure it matches the other matrices
rownames(precip.syrph.dis)<-precip_syrph$SiteYear
colnames(precip.syrph.dis)<-precip_syrph$SiteYear
```
## Now we use canopy cover from the NLCD (not the land cover raster, but their separate canopy cover one) (sort of as a proxy for area?) THis is just from 2021, but could extract from each year I think? 
```{r}
canopy<-read_csv("EnvironmentalData/canopy_2km.csv")%>%
  unique()%>%
  mutate(Site= substr(SiteYear, 1, 2))%>%
  select(-SiteYear)

canopy.bee<-left_join(bee.sites, canopy, by="Site")%>%
  unique()%>%
  arrange(Site) 
canopy.syrph<-left_join(syr.sites, canopy, by="Site")%>%
  unique()%>%
  arrange(Site) 

canopy.bee.dis<-dist(canopy.bee,method = "maximum",diag=FALSE)
canopy.bee.dis<-as.matrix(canopy.bee.dis)
rownames(canopy.bee.dis)<-precip_bee$SiteYear
colnames(canopy.bee.dis)<-rownames(canopy.bee.dis)

#same for syrphids
canopy.syrph.dis<-dist(canopy.syrph,method = "maximum",diag=FALSE)
canopy.syrph.dis<-as.matrix(canopy.syrph.dis)
rownames(canopy.syrph.dis)<-precip_syrph$SiteYear
colnames(canopy.syrph.dis)<-rownames(canopy.syrph.dis)
```
## Now for soil data
```{r}

```

